### 2022-07-14 draft.tex
Deploying to gh-pages from @ graciellehigino/ms_range_interactions@8668df84be3ce4d66689ced6284ae5641ea452e4 🚀

Finding a species in a certain location is like finding an encrypted
message that traveled through time. It carries the species' evolutionary
history, migration patterns, as well as any direct and indirect effects
generated by other species (some of which we may not even know exist).
Ecologists have been trying to decode this message with progressively
more powerful tools, from their field notes to highly complex
computational algorithms. However, to succeed in this challenge it is
important to have the right clues in hand. There are many ways we can be
misled by data - or the lack of it: taxonomic errors, geographic
inaccuracy, or sampling biases (Ladle and Hortal 2013; Hortal et al.
2015; Poisot et al. 2021). One way to identify - and potentially fix -
these errors is to combine many different pieces of information about
the occurrence of a species, so agreements and mismatches can emerge.
Here we suggest jointly analyzing species occurrence (range maps and
point occurrences) and ecological interactions to identify mismatches
between datasets.

Interactions form complex networks that shape ecological structures and
maintain the essential functions of ecosystems, such as seed dispersal,
pollination, and biological control (Albrecht 2018; Fricke et al. 2022)
that ultimately affect the composition, richness, and successional
patterns of communities across biomes. Yet, the connection between
occurrence and interaction data is a frequent debate in ecology
(Blanchet, Cazelles, and Gravel 2020). For instance, macroecological
models are often used with point or range occurrence data in order to
investigate the dynamics of a species with its environment. However,
these models do not account for ecological interactions, which might
largely affect species distribution (Abrego et al., n.d.; Afkhami,
McIntyre, and Strauss 2014; Araújo, Marcondes-Machado, and Costa 2014;
Godsoe et al. 2017; Godsoe and Harmon 2012). Some researchers argue that
occurrence data can also capture real-time interactions (Roy et al.
2016; Ryan et al. 2018), and, because of that, it would not be necessary
to include ecological interaction dynamics in macroecological models. On
the other hand, many mechanistic simulation models in ecology have
considered the effect of competition and facilitation in range shifts,
whilst the use of trophic interactions in this context remains
insufficient (Cabral, Valente, and Hartig 2017).

A significant challenge in this debate is the quality and quantity of
species distribution and ecological data (Boakes et al. 2010; Ronquillo
et al. 2020; Meyer, Weigelt, and Kreft 2016) - a gap that can lead to
erroneous conclusions in macroecological research (Hortal et al. 2008).
Amongst the geographical data available are the range maps provided by
the International Union for the Conservation of Nature (IUCN). Such maps
consist of simplified polygons, often created as alpha or convex hulls
around known species locations, refined by expert knowledge about the
species (2021). These maps can be used in macroecological inferences in
the lack of more precise information (Fourcade 2016; Alhajeri and
Fourcade 2019), but it has been recommended that they are used with
caution since they tend to underestimate the distribution of species
that are not well-known (Herkt, Skidmore, and Fahr 2017), do not
represent spatial variation in species occurrence and abundance (Dallas,
Pironon, and Santini 2020), and can include inadequate areas within the
estimated range. Another source of species distribution information is
the Global Biodiversity Information Facility (GBIF), which is an online
repository of georeferenced observational records that come from various
sources, including community science programs, museum collections, and
long-term monitoring schemes. A great source of bias in these datasets
is the irregular sampling effort, with more occurrences originating from
attractive and accessible areas and observation of charismatic species
(Alhajeri and Fourcade 2019). As for ecological data, a complete
assessment is difficult and is aggravated by biased sampling methods and
data aggregation (Poisot et al. 2020; Hortal et al. 2015). Nevertheless,
we have witnessed an increase in the availability of biodiversity data
in the last decades, including those collected through community science
projects (Callaghan et al. 2019; Pocock et al. 2015) and dedicated
databases, such as Mangal (Poisot et al. 2016). This provides an
opportunity to merge species distribution and ecological interaction
data to improve our predictions of where a species may be found across
large spatial scales (e.g., continental and global).

In this context, we elaborate a method that allows us to refine
distribution data (more precisely range maps) based on interaction data,
considering the basic assumption that predators can only be present in
regions where they are connected to at least one herbivore - and thus
indirectly connected to primary producers. We used a Serengeti food web
dataset (Baskerville et al. 2011) (which comprises carnivores,
herbivores, and plants from Tanzania) to demonstrate how a mismatch
between occurrence and interaction data can highlight significant
uncertainty areas in IUCN range maps. Finally, we add the GBIF
occurrence points for the Serengeti species to the investigation,
discuss the mechanisms that can lead to the lack of agreement between
data, and build from that a vision for the next steps, reinforcing the
importance of geographically explicit interaction data.

@@

Organisms cannot persist unless they are directly or indirectly
connected to a primary producer within their associated food web (Power
1992). Therefore, the range of a predator (omnivore or carnivore)
depends on the overlapping ranges of its preys. If sections of a
predator's range does not overlap with at least one of its prey it will
become disconnected from primary producers, and therefore we would not
expect the predator to occur in this area. This mismatch can be the
result of different mechanisms, like the overestimation of the
predator's range, taxonomic errors, or the lack of information about
trophic links. Thus, given that herbivores are the main connection
between plant resources (directly limited by environmental conditions)
and predators (Dobson 2009; Scott et al. 2018), here we adjusted the
ranges of predators based on a simple rule: we removed any part of a
predator's range that did not intersect with the range of at least one
prey herbivore species. So, unless the range of the predator overlapped
with at least one prey item, which in turn is directly connected to a
primary producer (plants), we removed that section of the predator's
range. Finally, we calculated the difference in range size between the
original IUCN ranges and those adjusted based on species interaction
data.

@@

There was high variation in the overlap of predator and prey ranges
(fig.~\ref{fig:geo_diss}). The high density of points on the left-hand
side of fig.~\ref{fig:geo_diss} indicates that most preys have small
ranges in comparison to those of the set of carnivores in the networks,
resulting in either low overlap between both ranges (bottom) or high
overlap of ranges because much of that of the prey is within predators'
range (top). The top-right side of the plot encompasses situations where
the ranges of both predator and prey are similar and overlapping, while
the bottom-right part of the plot represents a situation where the range
of the predator is smaller than that of its prey and much of it occurs
within the preys' range. For example, \emph{Panthera pardus} had many
preys occurring inside its range, with highly variable levels of overlap
(tbl.~\ref{tbl:everyone}). In general, species exhibited more consistent
values of prey-predator overlap, than predator-prey overlap -- indicated
by the spread of points along the x-axis, yet more restricted variation
on the y-axis (fig.~\ref{fig:geo_diss}). There was also no overall
relationship between the two metrics, or for any predator species.

@@

\caption{\label{tbl:everyone}List of species analyzed, their out and in

@@

Here we show that when ecological interaction data (predator-prey
interactions within food webs) are used to refine species range maps,
there are significant reductions in the IUCN range size of predatory
organisms. Despite showing the potential importance of accounting for
species interactions when estimating the range of a species, it remains
unclear the extent to which the patterns observed represent ecological
processes or a lack of data. In the following sections, we discuss the
implications of our findings, in terms of species range maps,
interaction data, and the next steps required to enhance understanding
of species distributions using information on ecological networks.

@@

Here we demonstrated how we can detect uncertainty in species
distribution data using ecological interactions. Knowing where
questionable occurrence data are can be crucial in ecological modelling
(Hortal 2008; Ladle and Hortal 2013), and accounting for these errors
can improve model outputs by diminishing the error propagation (Draper
1995). For instance, we believe this is a way to account for ecological
interactions in habitat suitability models without making the models
more complex, but making sure (not assuming) that the input data - the
species occurrence - actually accounts for ecological interactions. It
is important to notice, however, that the quality and usefulness of this
method are highly correlated with the amount and quality of data
available about species' occurrences \textbf{and} interactions. With
this paper we hope to add to the collective effort to decode the
encrypted message that is the occurrence of a species in space and time.
A promising avenue that adds to our method is the prediction of networks
and interactions in large scales (Strydom et al. 2021), for they can add
valuable information about ecological interactions where they are
missing. Additionally, in order to achieve a robust modelling framework
towards actual species distribution models we should invest in efforts
to collect and combine open data on species occurrence and interactions,
especially because we may be losing ecological interactions at least as
fast as we are losing species (Valiente-Banuet et al. 2015).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-AbregoAccSpe}{}%
Abrego, Nerea, Tomas Roslin, Tea Huotari, Yinqiu Ji, Niels Martin
Schmidt, Jiaxin Wang, Douglas W. Yu, and Otso Ovaskainen. n.d.
{``Accounting for Species Interactions Is Necessary for Predicting How
Arctic Arthropod Communities Respond to Climate Change.''}
\emph{Ecography} n/a (n/a). \url{https://doi.org/10.1111/ecog.05547}.

@@

@@

@@

\leavevmode\hypertarget{ref-IUCNSSCRedListTechnicalWorkingGroup2021MapSta}{}%
IUCN SSC Red List Technical Working Group. 2021. \emph{Mapping Standards
and Data Quality for IUCN Red List Spatial Data}.

@@

>>>>

Finding a species in a certain location is like finding an encrypted
message that traveled through time. It carries the species' evolutionary
history, migration patterns, as well as any direct and indirect effects
generated by other species (some of which we may not even know exist).
Ecologists have been trying to decode this message with progressively
more powerful tools, from their field notes to highly complex
computational algorithms. However, to succeed in this challenge it is
important to have the right clues in hand. There are many ways we can be
misled by data - or the lack of it: taxonomic errors (e.g., due to
updates in the taxonomy of a species), geographic inaccuracy (e.g.,
approximate coordinates or lack of documentation about their accuracy),
or sampling biases (e.g.~data clustered near roads or research centers)
(Ladle and Hortal 2013; Hortal et al. 2015; Poisot et al. 2021). One way
to identify - and potentially fix - these errors is to combine many
different pieces of information about the occurrence of a species, so
agreements and mismatches can emerge. Although previous studies have
combined different types of occurrence data to measure the accuracy of
datasets (Hurlbert and Jetz 2007; Hurlbert and White 2005; Ficetola et
al. 2014), none have used different types of information so far (i.e.,
ecological characteristics other than geographical distribution). Here
we suggest jointly analysing species occurrence (range maps and point
occurrences) and ecological interactions to identify mismatches between
datasets.

Interactions form complex networks that shape ecological structures and
maintain the essential functions of ecosystems, such as seed dispersal,
pollination, and biological control (Albrecht 2018; Fricke et al. 2022)
that ultimately affect the composition, richness, and successional
patterns of communities across biomes. Yet, the connection between
occurrence and interaction data is a frequent debate in ecology
{[}Blanchet, Cazelles, and Gravel (2020); Wisz2013RolBio{]}. For
instance, macroecological models are often used with point or range
occurrence data in order to investigate the dynamics of a species with
its environment. However, these models do not account for ecological
interactions, although it has been demonstrated that they might largely
affect species distribution {[}Abrego et al. (2021); Afkhami, McIntyre,
and Strauss (2014); Araújo, Marcondes-Machado, and Costa (2014); Godsoe
et al. (2017); Godsoe and Harmon (2012); Gotelli, Graves, and Rahbek
(2010); Wisz2013RolBio{]}. Some researchers argue that occurrence data
can also capture real-time interactions (see Roy et al. 2016; Ryan et
al. 2018), and, because of that, it would not be necessary to include
ecological interaction dynamics in macroecological models. On the other
hand, many mechanistic simulation models in ecology have considered the
effect of competition and facilitation in range shifts. For example,
Gotelli \emph{et al.} -(Gotelli, Graves, and Rahbek 2010) demonstrate
how conspecific attraction might be the main factor driving the
distribution of migratory birds; Afkhami \emph{et al.} -(Afkhami,
McIntyre, and Strauss 2014) explores how mutualistic fungal endophytes
are responsible or expanding the range of native grass; many other
examples are discussed in Wisz \emph{et al.} -(Wisz et al. 2013).
Although interactions across trophic levels are demonstrated to
determine species range (Wisz et al. 2013), the use of these
interactions in mechanistic simulation models in macroecology remains
insufficient (as discussed in Cabral, Valente, and Hartig 2017).

A significant challenge in this debate is the quality and quantity of
species distribution and ecological data (Boakes et al. 2010; Ronquillo
et al. 2020; Meyer, Weigelt, and Kreft 2016) - a gap that can lead to
erroneous conclusions in macroecological research (Hortal et al. 2008).
Amongst the geographical data available are the range maps provided by
the International Union for the Conservation of Nature (IUCN). Such maps
consist of simplified polygons, often created as alpha or convex hulls
around known species locations, refined by expert knowledge about the
species (IUCN Red List Technical Working Group 2019). These maps can be
used in macroecological inferences in the lack of more precise
information (Fourcade 2016; Alhajeri and Fourcade 2019), but it has been
recommended that they are used with caution since they tend to
underestimate the distribution of species that are not well-known
(Herkt, Skidmore, and Fahr 2017) (especially at fine scale resolutions;
(Hurlbert and Jetz 2007; Hurlbert and White 2005)), do not represent
spatial variation in species occurrence and abundance (Dallas, Pironon,
and Santini 2020), and can include inadequate areas within the estimated
range. Another source of species distribution information is the Global
Biodiversity Information Facility (GBIF), which is an online repository
of georeferenced observational records that come from various sources,
including community science programs, museum collections, and long-term
monitoring schemes. A great source of bias in these datasets is the
irregular sampling effort, with more occurrences originating from
attractive and accessible areas and observation of charismatic species
(Alhajeri and Fourcade 2019). As for ecological data, a complete
assessment is difficult and is aggravated by biased sampling methods and
data aggregation (Poisot et al. 2020; Hortal et al. 2015). Nevertheless,
we have witnessed an increase in the availability of biodiversity data
in the last decades, including those collected through community science
projects (Callaghan et al. 2019; Pocock et al. 2015) and dedicated
databases, such as Mangal (Poisot et al. 2016). This provides an
opportunity to merge species distribution and ecological interaction
data to improve our predictions of where a species may be found across
large spatial scales (e.g., continental and global).

It has been demonstrated that the agreement between range maps and point
data varies geographically (Hurlbert and Jetz 2007; Hurlbert and White
2005; Ficetola et al. 2014). Adding ecological interaction data to this
comparison might help to elucidate where these (dis)agreements are more
likely to be true and which dataset better represent the actual
distribution of species. In this context, we elaborate a method that
allows us to refine distribution data (more precisely range maps) based
on interaction data, considering the basic assumption that predators can
only be present in regions where they are connected to at least one
herbivore - and thus indirectly connected to primary producers. We used
a Serengeti food web dataset (Baskerville et al. 2011) (which comprises
carnivores, herbivores, and plants from Tanzania) to demonstrate how a
mismatch between occurrence and interaction data can highlight
significant uncertainty areas in IUCN range maps. Finally, we add the
GBIF occurrence points for the Serengeti species to the investigation,
discuss the mechanisms that can lead to the lack of agreement between
data, and build from that a vision for the next steps, reinforcing the
importance of geographically explicit interaction data.

@@

Organisms cannot persist unless they are directly or indirectly
connected to a primary producer within their associated food web (Power
1992). Therefore, the range of a predator (omnivore or carnivore)
depends on the overlapping ranges of its preys. If sections of a
predator's range does not overlap with at least one of its prey it will
become disconnected from primary producers, and therefore we would not
expect the predator to occur in this area. This mismatch can be the
result of different mechanisms, like the misestimation of both the
predator's and the preys' ranges {[}Ladle and Hortal (2013);
Rondinini2006TraDif{]}, taxonomic errors (Isaac, Mallet, and Mace 2004;
Ladle and Hortal 2013), or the lack of information about trophic links
(i.e., the lack of connection between the ranges of a predator and a
primary producer may be due a third species we don't know is connected
to both). Thus, given that herbivores are the main connection between
plant resources (directly limited by environmental conditions) and
predators (Dobson 2009; Scott et al. 2018), here we adjusted the ranges
of predators based on a simple rule: we removed any part of a predator's
range that did not intersect with the range of at least one prey
herbivore species. So, unless the range of the predator overlapped with
at least one prey item, which in turn is directly connected to a primary
producer (plants), we removed that section of the predator's range.
Finally, we calculated the difference in range size between the original
IUCN ranges and those adjusted based on species interaction data.

@@

There was a high variation in the overlap of predator and prey ranges
(fig.~\ref{fig:geo_diss}). The high density of points on the left-hand
side of fig.~\ref{fig:geo_diss} indicates that most preys have small
ranges in comparison to those of the set of carnivores in the networks,
resulting in either low overlap between both ranges (bottom) or high
overlap of ranges because much of that of the prey is within predators'
range (top). The top-right side of the plot encompasses situations where
the ranges of both predator and prey are similar and overlapping, while
the bottom-right part of the plot represents a situation where the range
of the predator is smaller than that of its prey and much of it occurs
within the preys' range. For example, \emph{Panthera pardus} had many
preys occurring inside its range, with highly variable levels of overlap
(tbl.~\ref{tbl:everyone}). In general, species exhibited more consistent
values of prey-predator overlap, than predator-prey overlap -- indicated
by the spread of points along the x-axis, yet more restricted variation
on the y-axis (fig.~\ref{fig:geo_diss}). There was also no overall
relationship between the two metrics, or for any predator species.

@@

\caption{\label{tbl:everyone}List of species analysed, their out and in

@@

Here we show that when ecological interaction data (predator-prey
interactions within food webs) are used to refine species range maps,
there are significant reductions in the IUCN range size of predatory
organisms. Despite showing the potential importance of accounting for
species interactions when estimating the range of a species, it remains
unclear the extent to which the patterns observed represent ecological
processes or a lack of data.

@@

Here we demonstrated how we can detect uncertainty in species
distribution data using ecological interactions. Knowing where
questionable occurrence data are can be crucial in ecological modelling
(Hortal 2008; Ladle and Hortal 2013), and accounting for these errors
can improve model outputs by diminishing the error propagation (Draper
1995). For instance, we believe this is a way to account for ecological
interactions in habitat suitability models without making the models
more complex, but by making sure (not assuming) that the input data -
the species occurrence - actually accounts for ecological interactions.
It is important to notice, however, that the quality and usefulness of
this method are highly correlated with the amount and quality of data
available about species' occurrences \textbf{and} interactions. With
this paper, we hope to add to the collective effort to decode the
encrypted message that is the occurrence of a species in space and time.
A promising avenue that adds to our method is the prediction of networks
and interactions in large scales (Strydom et al. 2021), for they can add
valuable information about ecological interactions where they are
missing. Additionally, in order to achieve a robust modelling framework
towards actual species distribution models we should invest in efforts
to collect and combine open data on species occurrence and interactions,
especially because we may be losing ecological interactions at least as
fast as we are losing species (Valiente-Banuet et al. 2015).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Abrego2021AccSpe}{}%
Abrego, Nerea, Tomas Roslin, Tea Huotari, Yinqiu Ji, Niels Martin
Schmidt, Jiaxin Wang, Douglas W. Yu, and Otso Ovaskainen. 2021.
{``Accounting for Species Interactions Is Necessary for Predicting How
Arctic Arthropod Communities Respond to Climate Change.''}
\emph{Ecography} 44 (6): 885--96.
\url{https://doi.org/10.1111/ecog.05547}.

@@

\leavevmode\hypertarget{ref-Ficetola2014EvaRob}{}%
Ficetola, Gentile Francesco, Carlo Rondinini, Anna Bonardi, Vineet
Katariya, Emilio Padoa-Schioppa, and Ariadne Angulo. 2014. {``An
Evaluation of the Robustness of Global Amphibian Range Maps.''}
\emph{Journal of Biogeography} 41 (2): 211--21.
\url{https://doi.org/10.1111/jbi.12206}.

@@

\leavevmode\hypertarget{ref-Gotelli2010MacSig}{}%
Gotelli, Nicholas J., Gary R. Graves, and Carsten Rahbek. 2010.
{``Macroecological Signals of Species Interactions in the Danish
Avifauna.''} \emph{Proceedings of the National Academy of Sciences} 107
(11): 5030--35. \url{https://doi.org/10.1073/pnas.0914089107}.

@@

\leavevmode\hypertarget{ref-Hurlbert2007SpeRic}{}%
Hurlbert, Allen H., and Walter Jetz. 2007. {``Species Richness,
Hotspots, and the Scale Dependence of Range Maps in Ecology and
Conservation.''} \emph{Proceedings of the National Academy of Sciences}
104 (33): 13384--89. \url{https://doi.org/10.1073/pnas.0704469104}.

\leavevmode\hypertarget{ref-Hurlbert2005DisRan}{}%
Hurlbert, Allen H., and Ethan P. White. 2005. {``Disparity Between Range
Map- and Survey-Based Analyses of Species Richness: Patterns, Processes
and Implications.''} \emph{Ecology Letters} 8 (3): 319--27.
\url{https://doi.org/10.1111/j.1461-0248.2005.00726.x}.

\leavevmode\hypertarget{ref-Isaac2004TaxInf}{}%
Isaac, Nick J. B., James Mallet, and Georgina M. Mace. 2004.
{``Taxonomic Inflation: Its Influence on Macroecology and
Conservation.''} \emph{Trends in Ecology \& Evolution} 19 (9): 464--69.
\url{https://doi.org/10.1016/j.tree.2004.06.004}.

\leavevmode\hypertarget{ref-IUCN2019MapSta}{}%
IUCN Red List Technical Working Group. 2019. {``Mapping Standards and
Data Quality for IUCN Red List Spatial Data.''} Prepared by the
Standards and Petitions Working Group of the IUCN SSC Red \ldots.

@@

\leavevmode\hypertarget{ref-Wisz2013RolBio}{}%
Wisz, Mary Susanne, Julien Pottier, W Daniel Kissling, Loïc Pellissier,
Jonathan Lenoir, Christian F Damgaard, Carsten F Dormann, et al. 2013.
{``The Role of Biotic Interactions in Shaping Distributions and Realised
Assemblages of Species: Implications for Species Distribution
Modelling.''} \emph{Biological Reviews of the Cambridge Philosophical
Society} 88 (1): 15--30.
\url{https://doi.org/10.1111/j.1469-185X.2012.00235.x}.
